Acceleration of Derivative Calculations with Application to Radial Basis Function -- Finite-Differences on the Intel MIC Architecture

### 2014-04-15 acm/ieee_paper_final.tex
adding author affiliations and replacing figures
typo

%\numberofauthors{4}

%\author{
%\alignauthor
%Gordon Erlebacher\\ %\titlenote{Dr.~Trovato insisted his name be first.}\\
%       \affaddr{Department of Scientific Computing}\\
%       \affaddr{Florida State University}\\
%       \affaddr{Tallahassee, FL 32306-4120}\\
%       \email{gordon.erlebach@gmail.com}
%\alignauthor
%Erik Saule
%        \affaddr{Department of Computer Science}\\
%        \affaddr{University of North Carolina at Charlotte}\\
%        \email{esaule@uncc.edu}
%\and
%\alignauthor
%Natasha Flyer
%        \affaddr{Computational and Information System Laboratory}\\
%        \affaddr{UCAR}\\
%        \email{flyer@ucar.edu}
%\alignauthor
%Evan Bollig
%        \affaddr{Minnesota Supercomputer Institute}\\
%        \affaddr{University of Minnesota}\\
%        \affaddr{Minneapolis, MN}\\
%        \email{bollig@gmail.com}
%}
\maketitle

@@

\begin{figure}[tbh]
\centering
\includegraphics[width=0.7\linewidth]{figures/RBFStencil_n32_a}\label{fig:stencil}
\caption{An example of an RBF-FD stencil of size $n=32$ on a sphere of $N=1024$ nodes to approximate any derivative operator at the point in the square. In nonsparse format, the DM  is $1024\times1024$ with 32 nonzero entries per row.}
\end{figure}

@@

%--------------------------------------------
\section{Modelization of the \\Potential Improvements}
\label{sec:model}

@@

\begin{figure}[t]
  \begin{center}
    \scalebox{.8}{
      \begin{tabular}{|c|l|}
        %\hline
        %& & \\
        \hline
        $b_i$ & number of bytes per index \\
        $b_x$ & number of bytes per value \\
        $n_z$ & number of nonzeros per row of $A$ \\
        $n_r$ & number of column/rows of $A$ \\
        $n_c$ & total number of nonzeros\\
        $n_v$ & number of {\tt x} vectors \\
        $n_m$ & number of matrices \\
        $s_M$ & size of the $n_m$ matrices in bytes\\
        $s_x$ & size of the $n_v$ {\tt x} vectors in bytes\\
        $s_y$ & size of the $n_v n_m$ {\tt y} vectors in bytes\\
        \hline
        $cl$    & size of a cache line in bytes\\
        $b_{wT}$ & number of bytes written to memory  \\
        $b_{rT}$ & minimum number of bytes read from memory  \\
        $b_T$   & minimum number of bytes transferred \\
        $B_{rT}$ & maximum number of bytes read from memory \\
        $B_T$   & maximum number of bytes transferred \\
        $O$     & number of floating point operations \\
        $I_b$   & maximum computational intensity\\
        $I_w$   & minimum computational intensity\\        
        \hline
      \end{tabular}
    }
  \end{center}
  \caption{Notation relative to the application problem (upper section) and to the benchmarks (lower section.)}
  \label{tab:not}
\end{figure}

@@

\begin{figure}
  \centering 
  \includegraphics[width=.9\linewidth]{figures/gflops_peak.pdf}\label{fig:gflops-peak-perf}
  \caption{Estimation of the maximum achievable performance using a
    device with a bandwidth from memory to computational units of 150~GB/s when
    varying the number of vectors and matrices.}
  \label{fig:perf_predict}
\end{figure}

@@

\ger{The main idea of this paper, computing 16 derivatives at a time,
can be applied on different architectures such as multiprocessor
multicore or GPU. To illustrate this fact, we conducted similar experiments
on Cascade at the Minnesota Supercomputing Institute
using two XEON E5 2670 nodes (8 cores, 2 threads and 20 Mbytes of L2
cache per core, with hyperthreading disabled.) The cores support the AVX instruction set. 
Both the compiler-generated code and the code written with the 
help of the AVX instruction set reach a read bandwidth of
85~GB/s. Using cache lines of 256 bits, we find a best/worst speeds of 4/20\; Gflops/s and 
50/120\; Gflops/s, respectively for the 1vec/1mat and 4vec/4mat cases. 
Our implementation on Cascade reaches 45 Gflops for the 4/4\, case, or 41\% of peak theoretical 
performance on this particular processor setup. For references, we achieve 
140/210=66\% of peak theoretical performance on the MIC. Note that peak performance 
is not that of the processor, but of the best possible algorithm on the processor. 
The gain in the 4/4 case with respect to 4x 1/1 case is respectively 4x on the MIC 
and 2.5x on the front end.  
}

>>>>

\numberofauthors{2}

\author{
\alignauthor Gordon Erlebacher\\ 
      \affaddr{Department of Scientific Computing}\\
      \affaddr{Florida State University}\\
%      \affaddr{Tallahassee, FL 32306-4120}\\
      \email{gordon.erlebach@gmail.com}
\alignauthor Erik Saule\\
       \affaddr{Department of Computer Science}\\
       \affaddr{University of North Carolina at Charlotte}\\
       \email{esaule@uncc.edu}
\and
\alignauthor Natasha Flyer\\
       \affaddr{Computational and Information Systems Laboratory}\\
       \affaddr{UCAR}\\
       \email{flyer@ucar.edu}
\alignauthor Evan Bollig\\
       \affaddr{Minnesota Supercomputer Institute}\\
       \affaddr{University of Minnesota}\\
%       \affaddr{Minneapolis, MN}\\
       \email{bollig@gmail.com}
}
\maketitle

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.65\linewidth]{figures/RBFStencil_n32_a}\label{fig:stencil}
\caption{An example of an RBF-FD stencil of size $n=32$ on a sphere of $N=1024$ nodes to approximate any derivative operator at the point in the square. In nonsparse format, the DM  is $1024\times1024$ with 32 nonzero entries per row.}
\end{figure}

@@

%--------------------------------------------
\section{Modeling the\\ Potential Improvements}
\label{sec:model}

@@

\begin{figure}[tbh]
  \begin{center}
    \scalebox{.8}{
      \begin{tabular}{|c|l|}
        %\hline
        %& & \\
        \hline
        $b_i$ & number of bytes per index \\
        $b_x$ & number of bytes per value \\
        $n_z$ & number of nonzeros per row of $A$ \\
        $n_r$ & number of column/rows of $A$ \\
        $n_c$ & total number of nonzeros\\
        $n_v$ & number of {\tt x} vectors \\
        $n_m$ & number of matrices \\
        $s_M$ & size of the $n_m$ matrices in bytes\\
        $s_x$ & size of the $n_v$ {\tt x} vectors in bytes\\
        $s_y$ & size of the $n_v n_m$ {\tt y} vectors in bytes\\
        \hline
        $cl$    & size of a cache line in bytes\\
        $b_{wT}$ & number of bytes written to memory  \\
        $b_{rT}$ & minimum number of bytes read from memory  \\
        $b_T$   & minimum number of bytes transferred \\
        $B_{rT}$ & maximum number of bytes read from memory \\
        $B_T$   & maximum number of bytes transferred \\
        $O$     & number of floating point operations \\
        $I_b$   & maximum computational intensity\\
        $I_w$   & minimum computational intensity\\        
        \hline
      \end{tabular}
    }
  \end{center}
  \caption{Notation relative to the application problem (upper section) and to the benchmarks (lower section.)}
  \label{tab:not}
\end{figure}

@@

\begin{figure}[t]
  \centering 
  \includegraphics[width=.9\linewidth]{figures/gflops_peak.pdf}\label{fig:gflops-peak-perf}
  \caption{Estimation of the maximum achievable performance using a
    device with a bandwidth from memory to computational units of 150~GB/s when
    varying the number of vectors and matrices.}
  \label{fig:perf_predict}
\end{figure}

@@

\ger{The main idea of this paper, computing 16 derivatives at a time,
can be applied on different architectures such as multiprocessor
multicore or GPU. To illustrate this fact, we conducted similar experiments
on Cascade at the Minnesota Supercomputing Institute
using two Intel Xeon E5 2670 nodes (8 cores, 2 threads and 20 Mbytes of L2
cache per core, with hyperthreading disabled.) The cores support the AVX instruction set. 
Both the compiler-generated code and the code written with the 
help of the AVX instruction set reach a read bandwidth of
85~GB/s. Using cache lines of 256 bits, we find a best/worst speeds of 4/20\; Gflops/s and 
50/120\; Gflops/s, respectively for the 1vec/1mat and 4vec/4mat cases. 
Our implementation on Cascade reaches 45 Gflops for the 4/4\, case, or 41\% of peak theoretical 
performance on this particular processor setup. For references, we achieve 
140/210=66\% of peak theoretical performance on the MIC. Note that peak performance 
is not that of the processor, but of the best possible algorithm on the processor. 
The gain in the 4/4 case with respect to 4x 1/1 case is respectively 4x on the MIC 
and 2.5x on the front end.  
}
